---
title: "Weight of ductular reaction"
author: "[Lluís Revilla](mailto:lrevilla@clinic.cat)"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
  BiocStyle::pdf_document:
    toc: true
---

```{r setup, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE, echo = TRUE, cache = TRUE, autodep = TRUE)
```

# Introduction

To evaluate the weight of the ductular reaction in the steps, we will perform a Gene Set Analysis using the top differentially expressed genes of the ductular reaction on the alcoholic liver disease progression and in the networks build.

To evaluate the relationship of the genes that increase or decrease the expression along the progression of disease with the ductular reaction, we will perform a GSEA on the top differentially expressed genes.

# Ductular reaction in co-expressed genes


```{r load_DR}
DR <- read.csv("DR_PosVsNeg.csv")
FC <- DR$logFC
names(FC) <- DR$Symbol
FC <- FC[DR$adj.P.Val < 0.05]
FC <- FC[!duplicated(names(FC))]
FC <- sort(FC, decreasing = TRUE)
pFC <- FC[DR$logFC > 0 & DR$adj.P.Val < 0.05]
nFC <- FC[DR$logFC < 0 & DR$adj.P.Val < 0.05]
```

## Early states of ALD

```{r}
library("limma")
```

## Late states of ALD

# Ductular reaction in progressive genes of ALD
This step is to evaluate the weight of the ductular reaction in the alcoholic liver disease progression. 

```{r load_kendall}
kendall <- read.csv("Kendall_ALD.csv")
K <- kendall$Tau
names(K) <- kendall$Symbol
K <- sort(K, decreasing = TRUE)
K <- K[kendall$p.value < 0.05]
index1 <- ids2indices(names(pFC), names(K))$Set1
index2 <- ids2indices(names(nFC), names(K))$Set1
barcodeplot(K, index = index1, index2 = index2, quantiles = c(-0.3, 0.3),
            labels = c("Increasing", "Decreasing"), 
            main = "Fold change in the progression")
index <- ids2indices(names(FC), names(K))$Set1
barcodeplot(K, index = index, labels = c("Increasing", "Decreasing"),
            quantiles = c(-0.3, 0.3))
```

## DR over-expressed in the progression of ALD

```{r}
barcodeplot(K, index = index1, labels = c("Increasing", "Decreasing"), 
            main = "Over-expressed genes of the DR in the progression", 
            col.bars = "red", quantiles = c(-0.3, 0.3))
```

## DR under-expressed in the progression of ALD

```{r}
barcodeplot(K, index = index2, labels = c("Increasing", "Decreasing"), 
            main = "Under expressed genes of the DR in the progression", 
            col.bars = "blue", quantiles = c(-0.3, 0.3))
```

# Progressive genes in the ductular reaction

```{r}
increasing <- K > 0.5
decreasing <- K < -0.5

index1 <- ids2indices(names(K[increasing]), names(FC))$Set1
index2 <- ids2indices(names(K[decreasing]), names(FC))$Set1
barcodeplot(FC, index = index1, index2 = index2, 
            main = "Progression in ductular reaction")
index <- ids2indices(names(K), names(FC))$Set1
barcodeplot(FC, index = index)
```


## Increasing expression with the progression of ALD
```{r}
barcodeplot(FC, index = index1, col.bars = "red", quantiles = c(-1.5, 1.5),
            main = "Increasingly expressed in the progression in the DR")
```


## Decreasing expression with the progression of ALD
```{r}
barcodeplot(FC, index = index2, col.bars = "blue", quantiles = c(-1.5, 1.5),
            main = "Decreasingly expressed in the progression in the DR")
```


# SessionInfo

```{r}
sessionInfo()
```

