---
title: "The weight of the ductular reaction in alcohol liver disease"
author: "[Lluís Revilla](mailto:lrevilla@clinic.cat)"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
  BiocStyle::pdf_document:
    toc: true
---

```{r setup, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE, echo = TRUE, cache = TRUE, autodep = TRUE)
```

# Steps

## Ductular reaction signature

Obtain the ductular reaction signature the top 2000 genes differentially genes

## Alcoholic liver disease signature

Obtain the alcoholic liver disease signature of the progression of the disease the top 2000 differentially expressed genes.

## Networks
Use [WGCNA](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/) to build the co-expressed genes of the alcoholic liver disease. And test the enrichment in the modules with genes differentially expressed genes of the ductular reaction.

## BioCor

Test if [BioCor](https://github.com/llrs/BioCor) works better finding, and grouping genes.

# Analysis

You can see the analysis of data in this folder:

 - [Ductular reaction](Ductular_reaction.html)
 - [Alcoholic disease](Alcoholic_disease.html)

# SessionInfo

```{r end}
sessionInfo()
```

