---
title: "Preparating data for the network creation with WGCNA"
author: "[Lluís Revilla](mailto:lrevilla@clinic.cat)"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
  BiocStyle::pdf_document:
    toc: true
---

```{r setup, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE, echo = TRUE, cache = TRUE, autodep = TRUE)
```

# Load data
We start from the expression data we previously normalized for the differential gene expression analysis.
```{r load, collapse = TRUE}
load(file = "Expression_ALD.RData", verbose = TRUE)
load("ALD.RData", verbose = TRUE)
library("WGCNA")
```
Of the second file we are only interested in the design2 object.

# Filtering
Firt we convert it to the format WGCNA requires (each row a sample, each column a gene). We
```{r transpose}
data.wgcna <- t(v1$E)
vclin0 <- design2[, "Progression", drop = FALSE]
```
We will make a network of the whole disease, another for the early steps from Normal to compensated cirrhosis and from non-severe alcoholic hepatitis to severe alcoholic hepatitis.
```{r filter}
early.patients <- phenoData$Status %in% c("ASH", "C.Comp", "Normal")
early.ALD <- data.wgcna[early.patients, ]
late.ALD <- data.wgcna[!early.patients, ]
```

# Export
For an easy use with the TFM repository that will be used I rename the objects and store them separately:
```{r export1}
vclin <- vclin0
gsg <- goodSamplesGenes(data.wgcna)
data.wgcna <- data.wgcna[gsg$goodSamples, gsg$goodGenes]
save(data.wgcna, vclin, file = "Whole_Network.RData")
```
We can check that the whole data cohort has `r nrow(data.wgcna)` samples and `r sum(gsg$goodGenes)` genes have non zero variance.
```{r export2}
gsg <- goodSamplesGenes(early.ALD)
data.wgcna <- early.ALD[gsg$goodSamples, gsg$goodGenes]
early.vclin <- vclin0[rownames(phenoData) %in% rownames(early.ALD), , drop = FALSE]
vclin <- early.vclin
save(data.wgcna, vclin, file = "Early_Network.RData")
```
The cohort of patients from the early state of the disease has `r nrow(data.wgcna)` samples and `r sum(gsg$goodGenes)` genes have non zero variance.
```{r export3}
gsg <- goodSamplesGenes(late.ALD)
data.wgcna <- late.ALD[gsg$goodSamples, gsg$goodGenes]
late.vclin <- vclin0[rownames(phenoData) %in% rownames(late.ALD), , drop = FALSE]
vclin <- late.vclin
save(data.wgcna, vclin, file = "Late_Network.RData")
```
So the remaining samples are `r nrow(data.wgcna)` patients and `r sum(gsg$goodGenes)` genes have non zero variance.

We have enough samples to build a robust network with WGCNA for each network. See the first [FAQ](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/faq.html) answer.
# Check data

In order to see if the data really reflects what the model we can plot it in a PCA:
```{r PCA, collapse = TRUE}
library("ggbiplot")
pca.graph <- function(celfiles=NULL, data=NULL, file, outcome = NULL,
                      col = NULL, ...) {
  # Data is the normalized, celfiles are the raw files
  if (is.null(data)) {
    if (!is.null(celfiles)) {
      data <- rma(celfiles)
    } else {
      stop("celfiles or data must be provided")
    }
  }

  if ("ExpressionSet" %in% is(data)) {
    if (is.null(outcome)) {
      outcome <- as.character(phenoData(data)$Type)
    }
    dists <- as.dist(1 - cor(exprs(data), method = "spearman"))
  } else {
    if (is.null(outcome)) {
      outcome <- rep("AH", ncol(data))
    }
    dists <- as.dist(1 - WGCNA::cor(data, method = "spearman",
                                    use = "pairwise.complete.obs"))
  }

  cmd <- cmdscale(dists, eig = TRUE, add = TRUE)
  perc.v <- round(cmd$eig/sum(cmd$eig)*100, 1)
  pdf(file)
  pca.plo <- ggbiplot(prcomp(dists, scale. = TRUE),
                      obs.scale = 1, var.scale = 1, ellipse = TRUE,
                      group = outcome, var.axes = FALSE,
                      # circle = TRUE
  )
  plot(pca.plo)
  plot(cmd$points[, 1], cmd$points[, 2], type = "n", main = "MDS samples",
       xlab = paste0("PC1 (", perc.v[1], "% explained var.)"),
       ylab = paste0("PC2 (", perc.v[2], "% explained var.)"),
       ...)
  text(cmd$points[,1 ], cmd$points[, 2], col = col, cex = 0.9,
       labels = outcome)
  dev.off()
  invisible(cmd)
}
MDS <- pca.graph(data = v1$E, file = "Whole_set.pdf", outcome = phenoData$Status)
pca.graph(data = t(early.ALD), file = "Early_ALD.pdf", outcome = early.vclin)
pca.graph(data = t(late.ALD), file = "Late_ALD.pdf", outcome = late.vclin)
rownames(MDS$points) <- phenoData$Status
plot(MDS$points, main = "Samples differences", col = c("black", "green", "yellow", "orange", "darkorange", "red", "red"))
legend("bottomleft", legend = unique(phenoData$Status), fill = c("black", "green", "yellow", "orange", "darkorange", "red"))
```

As we can see on the MDS (and others not shown) the data reflects the differences and confirms the model. Furthermore, it also confirms that dividing in this two steps also provides a good distinction between samples.

# Network building

There are many option to find the co-expressed genes with WGCNA, here I will explain some of the most important:

* networkType
  This option changes how are negative correlations (You can also use a distance function)considered when building the adjacency matrix . There are three options: 
    + "unsigned": This option consider correlations independently of the sign.
    + "signed": Considers as higher similarity those with positive correlation than with negative correlation
    + "signed hybrid": Considers only the positive correlations.
* TOMType
  This option changes how are topological overlap matrices (TOM) calculated. There are two options: 
    + "unsigned": Don't uses the sign of the correlations to build the TOM matrix from the adjacencies
    + "signed": Uses the sign of the correlations to build the TOM matrix from the adjacencies

Depending on what is the question one wants to answer should choose between one of those. 
# SessionInfo
Here are the packages and the versions used to analyse these data and build this page:
```{r end}
sessionInfo()
```
