---
title: "Preparating data for the network creation with WGCNA"
author: "[Lluís Revilla](mailto:lrevilla@clinic.cat)"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
  BiocStyle::pdf_document:
    toc: true
---

```{r setup, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE, echo = TRUE, cache = TRUE, autodep = TRUE)
```

# Introduction

This is the preparation of the datasets and data to use [WGCNA](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/) to build the co-expressed genes of the alcoholic liver disease. 
You can find the other steps of the study [here](index.html).

# Load data
We start from the expression data we previously normalized for the differential gene expression analysis.
```{r load, collapse = TRUE}
load(file = "Expression_ALD.RData", verbose = TRUE)
load("ALD.RData", verbose = TRUE)
library("WGCNA")
```
Of the second file we are only interested in the design2 object.

# Filtering
Firt we convert it to the format WGCNA requires (each row a sample, each column a gene).
```{r transpose}
data.wgcna <- t(v1$E)
vclin0 <- design2[, "Progression", drop = FALSE]
```
We will make several data sets:

 - To see if there is a group of genes co-expressed among all the samples a data set with all the data from, normal to severe alcoholic hepatitis.
 - To observe the co-expression in the early early steps of the disease we will make a data set from compensated cirrhosis and alcoholic steathosis heaptitis samples.
 - To observe the co-expression in the late steps of the alcohol liver disease we store a data set from non-severe alcoholic hepatitis to severe alcoholic hepatitis samples. 
 - To observe if there are co-expressed genes in the disease, another data set with all the patients from alcoholic steathosis to severe hepatitis.
 
```{r filter}
early.patients <- phenoData$Status %in% c("ASH", "C.Comp")
late.patients <- phenoData$Status %in% c("AH", "Responders", "Non-responders")
early.ALD <- data.wgcna[early.patients, ]
late.ALD <- data.wgcna[late.patients, ]
patients <- data.wgcna[late.patients | early.patients, ]
```

# Check data

In order to see if the data really reflects what the model we can plot it in a PCA:
```{r PCA, collapse = TRUE}
library("ggbiplot")

dists <- function(data) {
  as.dist(1 - WGCNA::cor(data, method = "spearman",
                                  use = "pairwise.complete.obs"))
}

MDS.graph <- function(dists, outcome, col=NULL) {
  
  cmd <- cmdscale(dists, eig = TRUE, add = TRUE)
  perc.v <- round(cmd$eig/sum(cmd$eig)*100, 1)
  plot(cmd$points[, 1], cmd$points[, 2], type = "n", main = "MDS",
       xlab = paste0("PC1 (", perc.v[1], "% explained var.)"),
       ylab = paste0("PC2 (", perc.v[2], "% explained var.)"))
  text(cmd$points[,1 ], cmd$points[, 2], col = col, cex = 0.9,
       labels = outcome)
  invisible(cmd)
}

PCA.graph <- function(dists, outcome) {
  pca <- prcomp(dists, scale. = TRUE)
  pca.plo <- ggbiplot(pca, obs.scale = 1, var.scale = 1, ellipse = TRUE,
                      group = outcome, var.axes = FALSE, main = "PCA")
  plot(pca.plo)
  invisible(pca)
}

explore <- function(data, outcome, col, file) {
  dists <- dists(data)
  pdf(file)
  PCA.graph(dists, outcome)
  MDS.graph(dists, outcome, col)
  dev.off()
}

colors <- c(rep("red", 18), rep("orange", 11), rep("yellow", 12), rep("pink", 9), rep("green", 10))
explore(data = v1$E, file = "Whole_set.pdf", outcome = phenoData$Status, 
        col = colors)

early.vclin <- vclin0[rownames(phenoData) %in% rownames(early.ALD), , drop = FALSE]
explore(data = t(early.ALD), file = "Early_ALD.pdf", 
        outcome = phenoData$Status[30:50], col = colors[30:50])

late.vclin <- vclin0[rownames(phenoData) %in% rownames(late.ALD), , drop = FALSE]
explore(data = t(late.ALD), file = "Late_ALD.pdf", outcome = phenoData$Status[1:29],
        col = colors[1:29])
patients.vclin <- late.vclin <- vclin0[rownames(phenoData) %in% rownames(late.ALD), , drop = FALSE]
explore(data = t(patients), file = "patients.pdf", outcome = phenoData$Status[1:50],
        col = colors[1:50])

d <- dists(v1$E)
PCA.graph(d, phenoData$Status)
```

As we can see on the MDS (and others not shown, but stored in the files) the data reflects the differences and confirms the model. Furthermore, it also confirms that dividing in this two steps also provides a good distinction between samples. However there is a clear outlier: the sample E49. It has clearly a different expression than other samples, maybe because she/he had other complications. In order to avoid introducing noise to the networks we remove this sample for the network constructions:

```{r outlier}
data.wgcna <- data.wgcna[!"E49" == rownames(data.wgcna), ]
patients <- patients[!"E49" == rownames(patients), ]
early.ALD <- early.ALD[!"E49" == rownames(early.ALD), ]
```
We can store the data in the right format for a study with WGCNA.

# Export
For an easy use with the TFM repository that will be used I rename the objects and store them separately:
```{r export1}
vclin <- vclin0
gsg <- goodSamplesGenes(data.wgcna)
data.wgcna <- data.wgcna[gsg$goodSamples, gsg$goodGenes]
save(data.wgcna, vclin, file = "Whole_Network.RData")
```
We can check that the whole data cohort has `r nrow(data.wgcna)` samples and `r sum(gsg$goodGenes)` genes have non zero variance. 
```{r export2}
gsg <- goodSamplesGenes(early.ALD)
data.wgcna <- early.ALD[gsg$goodSamples, gsg$goodGenes]
vclin <- early.vclin[!"E49" == rownames(early.vclin), , drop = FALSE]
save(data.wgcna, vclin, file = "Early_Network.RData")
```
The cohort of patients from the early state of the disease has `r nrow(data.wgcna)` samples and `r sum(gsg$goodGenes)` genes have non zero variance.
```{r export3}
gsg <- goodSamplesGenes(late.ALD)
data.wgcna <- late.ALD[gsg$goodSamples, gsg$goodGenes]
vclin <- late.vclin
save(data.wgcna, vclin, file = "Late_Network.RData")
```
So the remaining samples are `r nrow(data.wgcna)` patients and `r sum(gsg$goodGenes)` genes have non zero variance.

```{r export4}
gsg <- goodSamplesGenes(patients)
data.wgcna <- patients[gsg$goodSamples, gsg$goodGenes]
vclin <- patients.vclin[!"E49" == rownames(patients.vclin), , drop = FALSE]
save(data.wgcna, vclin, file = "patients.RData")
```

We have enough samples to build a robust network with WGCNA for each network. See the first [FAQ](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/faq.html) answer.

# Network building

There are many option to find the co-expressed genes with WGCNA, here I will explain some of the most important:

* networkType
  This option changes how are negative correlations (You can also use a distance function)considered when building the adjacency matrix . There are three options: 
    + "unsigned": This option consider correlations independently of the sign.
    + "signed": Considers as higher similarity those with positive correlation than with negative correlation
    + "signed hybrid": Considers only the positive correlations.
* TOMType
  This option changes how are topological overlap matrices (TOM) calculated. There are two options: 
    + "unsigned": Don't uses the sign of the correlations to build the TOM matrix from the adjacencies
    + "signed": Uses the sign of the correlations to build the TOM matrix from the adjacencies

Depending on what is the question one wants to answer should choose between one of those.

# SessionInfo
Here are the packages and the versions used to analyse these data and build this page:
```{r end}
sessionInfo()
```
